MultifidelityTrajectoryOptimization

### 2024-12-18 artefacts/paper.tex
plot analytical and statistical variance

\section{Problem Formulation}
This secition describes how the state space model of the dynamics is set up, along with the controller design and cost function for the trajectory optimization problem.
\subsection{Mass-Spring-Damper System}
For this paper, a simplified trajectory optimization problem is considered that is easy to understand and analyze.
The objective is to solve a reference tracking problem for the position and velocity of a simple mass-spring-damper system.
The mass-spring-damper system has the following parameters:
\begin{itemize}
  \item Mass: $m = 5$ g
  \item Spring constant: $k = 2$ N/m
  \item Damping coefficient: $c = 0.5$ Ns/m
    % \item Initial position: $x_0 = 0$ m
    % \item Initial velocity: $\dot{x}_0 = 0$ m/s
\end{itemize}

@@

\subsection{Stochastic LQR Controller}
We will consider the case where the initial state $x_0$ is stochastic, with mean $\mu$ and covariance $\Sigma$.
The control input vector $U$ is still deterministic.

@@

\section{Multi Level Monte Carlo}

This section formulates the stochastic trajectory optimization problem as a multi-level Monte Carlo problem.

@@

To determine if the low fidelity model can be used as a control variate, we need to determine if it coreelates with the high fidelity model.

System Setup:
\begin{itemize}
  \item High fidelity model/cost function: $J_h(u, x_0)$
  \item Low fidelity model/cost function: $J_l(u, x_0)$
  \item Control input in high fidelity model: $u_h$
  \item Size of $u_h$: $n_{u_h} \times 1$
  \item Control input in low fidelity model: $u_l$
  \item Size of $u_l$: $n_{u_l} \times 1$
  \item Random initial state: $x_0$
  \item High fidelity optimum: $u_h^*$
  \item Low fidelity optimum: $u_l^*$
\end{itemize}

@@

\begin{itemize}
  \item High fidelity model/cost function: $J_h$
  \item Low fidelity model/cost function: $J_l$
  \item High fidelity optimum: $u_h^*$
  \item High fidelity perturbation: $p_h$
  \item Perturbed high fidelity solution: $u_h = u_h^* + p_h$
  \item Perturbed high fidelity solution downsampled using restriction: $u_{hlr} = T_{hlr} u_h$
  \item Perturbed high fidelity solution downsampled using averaging: $u_{hla} = T_{hla} u_h$
\end{itemize}

@@

\begin{itemize}
  \item High fidelity model/cost function: $J_h$
  \item Low fidelity model/cost function: $J_l$
  \item Low fidelity optimum: $u_l^*$
  \item Low fidelity perturbation: $p_l$
  \item Perturbed low fidelity solution: $u_l = u_l^* + p_l$
  \item Perturbed high fidelity solution upsampled: $u_{lh} = T_{lh} u_l$
\end{itemize}

@@

\begin{itemize}
  \item High fidelity model/cost function: $J_h$
  \item Low fidelity model/cost function: $J_l$
  \item High fidelity solution: $u_h$
  \item High fidelity solution downsampled using restriction: $u_{hlr} = T_{hlr} u_h$
  \item High fidelity solution downsampled using averaging: $u_{hla} = T_{hla} u_h$
\end{itemize}

@@

\end{document}

>>>>

\section{Problem Formulation}
This secition describes how the state space model of the dynamics is set up, along with the controller design and cost function for the trajectory optimization problem.
\subsection{Mass-Spring-Damper System}
For this paper, a simplified trajectory optimization problem is considered that is easy to understand and analyze.
The objective is to solve a reference tracking problem for the position and velocity of a simple mass-spring-damper system.
The mass-spring-damper system has the following parameters:
\begin{itemize}
  \item $m = 5$ g : Mass
  \item $k = 2$ N/m : Spring constant
  \item $c = 0.5$ Ns/m : Damping coefficient
    % \item Initial position: $x_0 = 0$ m
    % \item Initial velocity: $\dot{x}_0 = 0$ m/s
\end{itemize}

@@

\subsection{Stochastic LQR Controller} \label{stochastic_lqr}
We will consider the case where the initial state $x_0$ is stochastic, with mean $\mu$ and covariance $\Sigma$.
The control input vector $U$ is still deterministic.

@@

\section{Monte Carlo with Control Variates}

This section formulates the stochastic trajectory optimization problem as a Monte Carlo problem with a low fidelity control variate.

@@

To determine if the low fidelity model can be used as a control variate, we need to determine if it correlates with the high fidelity model.

System Setup:
\begin{itemize}
  \item $J_h(u, x_0)$: High fidelity model/cost function
  \item $J_l(u, x_0)$: Low fidelity model/cost function
  \item $u_h$: Control input in high fidelity model, $n_{u_h} \times 1$
  \item $u_l$: Control input in low fidelity model, $n_{u_l} \times 1$
  \item $x_0$: Random initial state
  \item $u_h^*$: High fidelity optimum
  \item $u_l^*$: Low fidelity optimum
\end{itemize}

@@

\begin{itemize}
  \item $J_h$: High fidelity model/cost function
  \item $J_l$: Low fidelity model/cost function
  \item $u_h^*$: High fidelity optimum
  \item $p_h$: High fidelity perturbation
  \item $u_h = u_h^* + p_h$: Perturbed high fidelity solution
  \item $u_{hlr} = T_{hlr} u_h$: Perturbed high fidelity solution downsampled using restriction:
  \item $u_{hla} = T_{hla} u_h$: Perturbed high fidelity solution downsampled using averaging
\end{itemize}

@@

\begin{itemize}
  \item $J_h$: High fidelity model/cost function
  \item $J_l$: Low fidelity model/cost function
  \item $u_l^*$: Low fidelity optimum
  \item $p_l$: Low fidelity perturbation
  \item $u_l = u_l^* + p_l$: Perturbed low fidelity solution
  \item $u_{lh} = T_{lh} u_l$: Perturbed high fidelity solution upsampled
\end{itemize}

@@

\begin{itemize}
  \item $J_h$: High fidelity model/cost function
  \item $J_l$: Low fidelity model/cost function
  \item $u_h$: High fidelity solution
  \item $u_{hlr} = T_{hlr} u_h$: High fidelity solution downsampled using restriction
  \item $u_{hla} = T_{hla} u_h$: High fidelity solution downsampled using averaging
\end{itemize}

@@

It is also possible to compute the correlation between $J_h(u_h, x_0)$ and $J_l(u_{hlr}, x_0)$ for every iteration of the numerical optimizer.
This helps us determine if there is a certain iteration of $J_l(u_{hlr}, x_0)$ that correlates well with all iterations of $J_h(u_h, x_0)$.
The same comparison can be made with $J_l(u_{hla}, x_0)$.

Comparing correlation between $J_h(u_h, x_0)$ and $J_h(u_{hlr}, x_0)$ across all iterations

Comparing correlation between $J_h(u_h, x_0)$ and $J_l(u_{hla}, x_0)$ across all iterations

\subsection{Control Variate Construction}

The control variate is constructed using the low fidelity model/cost function, since we have found that there is some correlation between $J_h(u_h, x_0)$ and $J_l(u_{hla}, x_0)$.
Here we are only using the averaged downsampled high fidelity solution $u_{hla}$ as the control variate.
The important variables for constructing the control variate are:

\begin{itemize}
  \item $x_0$: Random initial state
  \item $J_h$: High fidelity model/cost function
  \item $J_l$: Low fidelity model/cost function
  \item $u_h$: High fidelity solution
  \item $u_{hla} = T_{hla} u_h$: High fidelity solution downsampled using averaging
  \item $S_n^{MC}(J_h(u_h, x_0))$: Monte Carlo estimate of the high fidelity cost function
  \item $S_n^{MC}(J_l(u_{hla}, x_0))$: Monte Carlo estimate of the low fidelity cost function
  \item $\alpha$: Control variate coefficient
\end{itemize}

The control variate coefficient $\alpha$ can be calculated as

\begin{equation}
  \alpha = -\frac{\text{Cov}(J_h(u_h, x_0), J_l(u_{hla}, x_0))}{\text{Var}[J_l(u_{hla}, x_0)]}
\end{equation}

Hence, the control variate is

\begin{equation}
  S_n^{CV}(J_h, J_l) = S_n^{MC}(J_h(u_h, x_0)) + \alpha (S_n^{MC}(J_l(u_{hla}, x_0)) - \mathbb{E}[J_l(u_{hla}, x_0)])
\end{equation}

Where the samples of $x_0$ used in $J_h(u_h, x_0)$ and $J_l(u_{hla}, x_0)$ are the same.

Note that $\text{Cov}(J_h(u_h, x_0), J_l(u_{hla}, x_0))$, $\text{Var}[J_l(u_{hla}, x_0)]$ and $\mathbb{E}[J_l(u_{hla}, x_0)]$ are calculated analytically using the equations in section \ref{stochastic_lqr}.

\subsection{Numerical optimizer using Control Variate}

The numerical optimizer was run in 3 different scenarios

\begin{itemize}
  \item $S_n^{MC}(J_h(u_h, x_0))$: MC estimate of the high fidelity cost function
  \item $S_n^{CV}(J_h, J_l)$: CV estimate with $u_{hla}$ at the current iteration
  \item $S_n^{CV_{max}}(J_h, J_l)$:CV estimate with $u_{hla}$ at the iteration with the highest correlation
\end{itemize}

For these experiments, $n$, the number of samples of $x_0$ used in the MC estimate, was compared between 10, 100, 500 and 1000.
The comparison is done between $S_n^{MC}(J_h(u_h, x_0))$ vs $S_{\frac{n}{2}}^{CV}(J_h, J_l)$ and $S_{\frac{n}{2}}^{CV_{max}}(J_h, J_l)$ to keep the number of cost function evaluations the same with the control variate estimators having a lower computational cost due to half the sammples being used in the low fidelity cost function.

To compare the variance, we ran a numerical optimizer with n random samples of $x_0$ and calculated the variance for each iteration.

The variance was also calculated analytically for each control variate
\end{document}
